---
title: "Digitizing protocol"
author: "Jeffrey A. Walker"
date: "11/5/2020"
output:
  html_document:
    number_sections: true
    theme: lumen
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(here)
image_folder <- c("static/research/zebrafish/img")
image_path_1 <- "/research/zebrafish/images/set_scale_menu_item.png"
```

**Download Files**

@. Download your files to your hardrive (this may change if I can get a true shared drive)

**Setup**

@. Open the project spreadsheet. It is a google sheet in the folder with the video files
@. Copy the name of the file you are going to open and paste into the project spreadsheet in the “file” field
@. In Fiji, open the video file using Import > AVI, choose OK (do not open by double clicking the video file itself)
@. Expand the size of the window containing the video to fill most of your screen and zoom in so the video fills the window size

**NOTE -- Do not click the File Menu and choose Save**. This is not how the digitized points are saved (see below).
    
**Set the scale**

@. Click on the line tool
@. Shift-click on two *distant* points along a grid-line in horizontal direction
@. Count the number of squares along this line. Each square is 1 cm so the distance between the points is the number of squares you counted
@. [Go to menu item Analyze > Set Scale](https://www.middleprofessor.com/research/zebrafish/images/set_scale_menu_item.png)
@. [Insert the distance in the “Know Distance” box](https://www.middleprofessor.com/research/zebrafish/images/set_scale_window.png) (2nd from top)
@. [Look at the Scale at the bottom. It will say something like “29.006 pixels/unit”.](https://www.middleprofessor.com/research/zebrafish/images/set_scale_window.png) This is the scale.
@. Insert this number (not the units!) into the “scale” field of the project spreadsheet.
    
**Digitize**

@. Find F0, the frame prior to the initial head movement. Write this frame number in the “F0” field of the project spreadsheet
@. Start one frame prior to F0 -- so F0 is the second digitized frame
@. Digitize 37 frames total (35 of movement + 2 initial frames)
@. Save file using file icon on point picker tool bar -- do not save using the File Menu!
    
**What to digitize**

@. [Digitize 10-12 points from the tip of the snout to the end of the vertebral column](https://www.middleprofessor.com/research/zebrafish/images/snout-tail.png)
@. The points have to be **in order** from snout to tail -- don't insert a point in between other points
@. [The points should be along the **midline** 1/2way between each lateral edge of the fish](https://www.middleprofessor.com/research/zebrafish/images/points.png)

**Upload Files**

@. Upload your text files to shared drive (this may change if I can get a true shared drive)

**Pointpicker**

* [the using pointpicker manual is here.](http://bigwww.epfl.ch/thevenaz/pointpicker/){target="_blank"}

```{r echo=FALSE, out.width="50%"}
image_path <- "images/point-picker-tool-bar.png"
knitr::include_graphics(image_path)
```


